Rethinking how students learn to design programs

### 2016-04-23 manuscript-body.tex
Copy code snapshotting intro from dissertation

\subsection{We need more research on how software gets designed}\label{we-need-more-research-on-how-software-gets-designed}

In 2010, the journal \emph{Design Studies} devoted a special issue to studies of how professional software engineers design complex systems \cite{petre_editorial_2010}. The journal issue featured five different research perspectives on the same dataset: videos of three professional software engineering teams trying to design a traffic simulator. The existence of the journal issue, the monograph that followed it \cite{petre_software_2014} and the investigations contained therein, were motivated by what the issue's editors saw as a pressing need:

\begin{quote}
\emph{Not enough is known about the formative stages of software design.} The software engineering community has produced numerous design tools and languages, but, in practice, when software designers are first presented with a novel design problem, they more often than not will eschew these tools and languages. During formative design, software engineers spend a great deal of time engaging in creative, exploratory design thinking using pen and paper or a whiteboard---whether alone or in a small group. However, not enough is known about how software designers work in such settings. What do designers actually do during early software design? How do they communicate? What sorts of drawings do they create? \emph{What kinds of strategies do they software engineers apply in exploring the vast space of possible designs?} \cite{petre_editorial_2010} p.~533; emphasis mine
\end{quote}

Research in both the special issue \cite{petre_editorial_2010} and the monograph \cite{petre_software_2014} takes on a variety of challenges in the study of expert software design practice. One challenge, for example, is that of understanding how engineers process, prioritize, and cope with design requirements. Ball, Linden, Onarheim, and Christensen \cite{ball_design_2010} observed that engineers deploy mixed strategies, developing solutions breadth-first for easy problems and depth-first for more complex problems. They also found that the more complex a requirement became, the more likely engineers were to create speculative simulations (through talk and representations) about how a system might work to solve that problem \cite{ball_design_2010}.

Looking at the design sessions longitudinally, Baker and van der Hoek \cite{baker_ideas_2010} explored the shape and trajectory of how ideas generated in the design process develop and relate to one another. Those researchers found roughly a third of the ideas discussed in a typical design session ``were reiterations or rephrasings of previously stated ideas'' \cite{baker_ideas_2010}. While it is perhaps frustrating that ideas would be repeated so much, the authors argue such repetition can be viewed as a kind of continual revisitation to make sure a proposed design coheres:

\begin{quote}
Rather than representing a failure on the part of the designers, this repetition seems to be a necessary character of successful design sessions. Each time an idea is resurrected it is placed them sic in a new context, and compared to different aspects of the system. In this way, a concept of compatible, elegant design ideas is slowly converged upon. \cite{baker_ideas_2010}
\end{quote}

@@

>>>>

A growing trend in computer science education research is the collection and analysis of code snapshot data—records of the state of and changes to students’ code as they develop it \cite{jadud_methods_2006,rodrigo_analyzing_2009,spacco_marmoset_2006}. Though specific implementations differ, the general strategy in such projects is that a student event (typically compiling code or saving a file) triggers a procedure that creates a record containing the entire content of all of a student's relevant files, as well as associated metadata (time of save/compilation, for example, and any compiler errors that may have been generated). Mining data from such snapshotting systems has led to large-scale documentation of common student errors \cite{spacco_marmoset_2006}, the development of compile-time detectors to catch common student errors \cite{spacco_software_2005}, and the proposal of formative assessment models to predict student success \cite{jadud_methods_2006,tabanao_predicting_2011}. Building on existing momentum, some researchers are actively pushing for a continued scale-up of how we collect code snapshot data. One current proposal even calls for creating an international database by collecting snapshot data from thousands of introductory programming students worldwide \cite{kolling_building_2012}.

What's common to these threads of research is how they mine the data. In
most applications of code snapshot\footnote{What I call ``code
  snapshot'' research may be more formally called ``online protocol''
  research \cite{jadud_methods_2006,rodrigo_analyzing_2009}. ``Online'' here isn't mean to mean
  the sense of globally-connected computers, but rather that student
  materials are collected in a minimally-intrusive way \emph{while
  students work}.} research, the aim is to average across events and
sessions to arrive at a characteristic measure of a student's behavior.
Jadud \cite{jadud_methods_2006}, for example, develops the idea of a student's ``Error
Quotient,'' a measure of how effectively a student addresses
compile-time errors in their code.\footnote{A useful way of thinking
  about Jadud's \cite{jadud_methods_2006} error quotient is that the computation believes
  in second chances. It doesn't penalize students for making errors; it
  penalizes students if they don't fix a given error before compiling
  the same piece of code \emph{again.}} Rodrigo and colleagues extend
the idea of error quotient and introduce similar
measures such as students' ``compilation profiles'' \cite{rodrigo_analyzing_2009}, ``frustration profiles'' \cite{rodrigo_affective_2009}, and ``error
profiles'' \cite{tabanao_predicting_2011}. Because these measures can be
collected in real-time as students code, they offer the potential to
identify at-risk students and respond with early
interventions.

But, for all the data these methods collect their focus is primarily
what Jadud (2006) calls students' ``syntactic'' struggles: the challenge
of articulating well-formed statements a compiler can properly parse.
This focus on syntax-level struggles leads to both a methodological and
theoretical trade-off: we can see in detail how students struggle with
wording, but we see less of how they struggle with the meaning and
intent of their code. Syntactical analyses necessarily ignore the
specific \emph{content} of students' code because they abstract
meta-information about the event: error type, error location, frequency
of error.

As an analogy, suppose that instead of studying computer science
students we were studying screenwriting students. Further suppose we
have a system to track data whenever students save their screenplays and
scripts. For each save, we get a copy of the entire script at that time.
We can also run an automated analysis to check whether they've properly
formatted slug lines\footnote{In a screenplay, a slug line gives
  information about where and when a scene takes place. For example:
  ``INT. FORTRESS OF SOLITUDE -- DAY.''}, put character names in
all-capital letters, properly numbered scenes, etc. We might even know
where students are when they write (coffee shop, library, home, etc.).
If we mine that data, we could learn something about students'
\emph{screenwriting behavior}, but only at the level of how they
struggle with screenwriting syntax. Using only error-based data it's
much harder to answer questions such as:

\begin{itemize}
\item
  How does this student construct a scene or handle dialog in their
  script?
\item
  Does the student seem to have a grasp of pacing, story beats, and
  efficient exposition?
\item
  How does their writing manage and develop character arcs?
\end{itemize}

We also still can't answer fundamental questions about the context of
students' writing processes:

\begin{itemize}
\item
  Do they use particular techniques to ``break story'' and decompose a
  narrative into its key beats (index cards? Whiteboard?)
\item
  Do they participate in a writing group?
\item
  How do they respond to people giving them notes on revising a script?
\end{itemize}

To sum up, because snapshot systems are designed to collect the totality
of a codebase at frequent intervals, they offer a rich record of data
that captures the results---both major and minor---of students' design
decisions. But, computing education research that \emph{uses}
code-snapshotting has focused much more on detecting, classifying, and
predicting student errors than it has on showing how students progress
in programming and design expertise. Nevertheless, snapshot-based
research shows tremendous promise. Given that :

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Snapshotting students' code represents a cutting-edge way to resolve
  the way code---as a design artifact---evolves over time, \emph{but}
\item
  Code snapshots, as they're currently used, explore neither the
  totality of a student's design nor the rich context of that design's
  production, \emph{and}
\item
  There is a lack of parity between studies of how professionals design
  software and studies of how students do so,
\end{enumerate}

It seems sensible to ask: \emph{can code snapshots be used---possibly in
synthesis with ethnographically-oriented methods---to start studying how
students' design thinking plays a role in their introductory programming
work?} We believe the answer is yes.

In what follows, we present work that proceeds from an empirical
challenge: how can we develop accounts of students' programming activity
that explain the form and evolution of their code on a design project?
We offer an account of one student---Rebecca---and her experiences and
code from a second-semester course on programming concepts for
engineers. Using data from both code snapshots and clinical interviews,
we explicate both the challenges of studying students' software design
processes and the potential for such study to inform accounts of
teaching and learning.

@@

\subsection{Acknowledgments}\label{acknowledgments}

Blinded for review.

\clearpage

\bibliographystyle{acm}
\bibliography{bibliography}
